# tibbles, ggplot2, and the _tidyverse_ {#ch:intro}

The tidyverse universe includes:

```{r, echo=FALSE}
library(tidyverse)
```

In general, the tidyverse is the following:

1. provided the `pipe` command `%>%`
  * `x %>% f(y, z, ...)` is `f(x, y, z, ...)`
  * allows chained commands for better coherence
  * e.g., `mtcars %>% apply(2, mean)` is error without `tidyr::%>%`
2. `tibble` is the improved data structure of the tidyverse
  * easier to read-in data to a useful format
  * automatic type conversion
  * nicer printing options
3. `dplyr` provides tibble manipulation commands
  * understandable data processing with `pipe` streams
  * **filter** data faster
  * **arrange** rows of data easily
  * **select** columns quickly
  * **mutate** variables
  * **summarize** according to `group_by()`
  * also provides SQL relational operations
4. `ggplot2` is a plotting syntax (grammar of graphics)
  * `qplot()` provides a sensible **q**uick **plot**
  * apply plot types to data rather than the reverse
  * e.g. `ggplot(data) + plot_type(aes(xvar, yvar, groups), options)`
  * allows grid of plots by group using **facets**
  * overlays statistical summaries, e.g. `+ geom_smooth(x, y)`
  * "add" options such as transformed axes, labels, coordinates, etc.
5. `readr` is a faster, less painful read-in method
  * `read_fun` denotes `readr` functions (instead of `read.fun`)
  * guesses column types
  * offers writing functions, too
  * allows read and write with RDS, R's binary format
6. `tidyr` recharacterizes tibbles
  * `spread()` turns key and value columns into key-category columns
  * e.g., `state, year, pop` into `state, 1990, 1991, ...` of pop values
  * `gather()` turns expands data frames by condensing columns
  * e.g., condenses `1990, 1991, ...` into two `year, pop` columns
7. Other helpful tidyverse packages:
  * `stringr` offers many useful `str_fun` operations
  * `forcats` has operations _for cat_egorical variables
  * `lubridate` provides date and time control
  * `purrr`

The examples I'll use in the next few chapters
are the Boston housing database
and the Lahman baseball database.
By doing analysis on these two data sets, I hope to
introduce the power of the tidyverse.

## Tibbles: Boston housing data
Load, convert, print a tibble.
```{r load-boston}
# Convert to a tibble so it prints nicely
library(MASS)
select <- dplyr::select
boston <- as_tibble(MASS::Boston)
boston
```

```{r boston-vars}
?MASS::Boston
```

* crim
per capita crime rate by town.
* zn
proportion of residential land zoned for lots over 25,000 sq.ft.
* indus
proportion of non-retail business acres per town.
* chas
Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).
* nox
nitrogen oxides concentration (parts per 10 million).
* rm
average number of rooms per dwelling.
* age
proportion of owner-occupied units built prior to 1940.
* dis
weighted mean of distances to five Boston employment centres.
* rad
index of accessibility to radial highways.
* tax
full-value property-tax rate per \$10,000.
* ptratio
pupil-teacher ratio by town.
* black
$1000(Bk - 0.63)^2$ where Bk is the proportion of blacks by town.
* lstat lower status of the population (percent).
* medv median value of owner-occupied homes in \$1000s.

A ggplot is the first declaration (usually variable `data` is defined),
followed by graphics definitions (operations on the data):
```{r price-rooms-1}
ggplot(data = boston) + 
  geom_point(mapping = aes(x = rm, y = medv), alpha=0.4) +
  labs(x = "avg. rooms per house",
       y = "median house value",
       title = "House values vs. size in Boston")
```

Making a histogram of all numeric variables. First step, gather all variables.
```{r intro-gather}
boston %>%
  keep(is.numeric) %>%  # strips all non-numeric columns (unnecessary here)
  gather()  # puts all variable values in a single column 'value'
```

Facet wrap allows plotting each `key` level separately.
```{r all-var-hist}
boston %>% 
  gather() %>%
  ggplot() +
    facet_wrap(~ key, scales = "free") +
    geom_histogram(mapping = aes(value), bins=20)
```

From the histograms, there seems to be only a few values of `crim` over 30.
```{r high-crime-obs}
boston %>%
  filter(crim > 30)
```

## ggplot2 and EDA
But we want to know the conditional distributions according to `medv`.
First, showing this with conditional densities.
```{r cond-hist}
boston %>% 
  gather('key', 'value', -medv) %>%
  mutate(price_gr = ntile(medv, 4)) %>%
  ggplot(aes(value, group = price_gr)) +
    facet_wrap(~ key, ncol = 4, scales = "free") +
    geom_freqpoly(aes(color = price_gr), bins = 12)
# Click on the expand icon at the top right to make bigger.
```

Appears `chas` is categorical.
```{r change-chas}
boston <- boston %>%
  mutate(chas = factor(chas))
```

Second, scatterplots of median value vs. all variables.
```{r medv-scatter, warning=F}
boston %>%
  gather('key', 'value', -c(medv, chas)) %>% 
  ggplot(aes(x = value, y = medv)) +
    facet_wrap( ~ key, scales = "free") +
    geom_point(aes(shape = chas), size = 0.5, alpha = 0.25) +
    geom_smooth(lwd = 0.5, se = TRUE) +
  ggsave('plots/medv-scatter.pdf')
# Click on the expand icon at the top right to make bigger.
```

There are ggplot jitter options, but none worked for me.
```{r medv-jitter-nw, warning=F}
boston %>%
  gather('key', 'value', -c(medv, chas)) %>% 
  ggplot(aes(x = value, y = medv)) +
    facet_wrap( ~ key, scales = "free") +
    geom_jitter(aes(shape = chas), size = 0.5, alpha = 0.25) + 
    geom_smooth(lwd = 0.5, se = TRUE)
```

Tinkering to get a jittered plot.
```{r medv-jitter}
var_sd <- boston %>%
  gather('key', 'value', -c(medv, chas)) %>% 
  group_by(key) %>% 
  summarize(var_sd = sd(value))
boston %>%
  gather('key', 'value', -one_of(c("medv", "chas"))) %>% 
  left_join(y = var_sd, by = "key") %>% 
  mutate(jit_val = value + var_sd * runif(nrow(boston), -0.1, 0.1)) %>% 
  ggplot(aes(x = jit_val, y = medv)) +
    facet_wrap( ~ key, scales = "free") +
    geom_jitter(aes(shape = chas), size = 0.5, alpha = 0.25) +
    geom_smooth(lwd = 0.5, se = TRUE) +
  ggsave('plots/medv-jitter.pdf')
```

Covariance plot of variables.
```{r cov-plot}
boston %>% 
  keep(is.numeric) %>% 
  cor() %>% 
  as_tibble() %>% 
  mutate(name = colnames(boston[sapply(boston, is.numeric)])) %>% 
  gather( , , -one_of("name")) %>% 
  ggplot(aes(name, key, fill = value)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                         breaks = seq(-1, 1, , by = 0.2)) +
    theme(legend.key.height = unit(45, "pt"))
```
But a better correlation plot is in a package designed for them.
```{r using-corrplot}
library(corrplot)
boston %>% 
  keep(is.numeric) %>% 
  cor() %>%
  abs() %>% 
  corrplot(cl.lim = c(0, 1))
```

Analyze median value and highway access `rad`.
```{r medv-rad}
boston %>%
  ggplot(aes(rad, medv)) +
  geom_jitter(aes(color = chas), 
              height = 2, width = NULL, 
              size = 1, alpha = 0.4) +
  geom_smooth(aes(color = chas), lwd = 1)
```

Perhaps `rad = 24` is a missing value.
```{r list-rad-val}
boston %>%
  count(rad)
```
```{r explore-rad}
boston %>%
  gather( , , -rad) %>% 
  group_by(key, rad) %>%
  mutate(value = as.numeric(value)) %>%  # necessary due to factor variable chas
  summarize(z = round(mean(value), 1)) %>% 
  spread(rad, z)
```

Or in helpful boxplot format.
```{r boxplot-rad}
boston %>%
  keep(is.numeric) %>% 
  gather( , , -rad) %>% 
  group_by(key, rad) %>%
  ggplot(aes(x = rad, y = value, group = rad)) + 
    geom_boxplot(outlier.size = 0.5, varwidth = T) +
    facet_wrap(~ key, ncol = 3, scales = "free") +
  ggsave('plots/rad-boxplot.pdf')
```

Looking at `lstat` relationships.
```{r violin-lstat}
boston %>%
  keep(is.numeric) %>% 
  gather( , , -lstat) %>% 
  mutate(lstat_gr = ntile(lstat, 10)) %>% 
  group_by(key, lstat_gr) %>%
  ggplot(aes(x = lstat_gr, y = value, group = lstat_gr)) + 
    geom_violin() +
    facet_wrap(~ key, ncol = 3, scales = "free") +
  ggsave('plots/lstat-violin.pdf')
```

Jittering works well for single plots.
```{r tax-nox}
boston %>%
  ggplot(aes(tax, nox)) +
    geom_jitter(aes(color = medv, shape = chas), 
                height = 0.02, width = 10) +
    scale_color_gradient2(midpoint = 20, 
                          low = "blue", mid = "gray75", high = "red") +
    geom_hline(yintercept = 0.6, color = "yellow") +
    geom_abline(slope = 0.001, intercept = 0.1, color = "blue", lty = "93133313")
  ggsave('plots/tax-nox.pdf')
```

## Many plotting options
Statistics can be added to the plot as an additional layer.
Other layers are coordinates, facets, and scales.
```{r price-rooms-2}
ggplot(data = boston) + 
  geom_point(mapping = aes(x = rm, y = medv, color = crim), alpha=0.75) +
  geom_smooth(mapping = aes(x = rm, y = medv)) +
  coord_cartesian(xlim = c(4.5, 7.5)) +
  scale_y_log10() +
  scale_color_gradient(low = "yellow", high = "blue") +
  labs(x = "average rooms / house", y = "median house price ($K)",
       title = "Boston median house prices vs. average house size")
```

Maybe more useful if colored by quantile of `crim` value.
```{r price-rooms-3}
boston %>% 
  mutate(crim = cume_dist(crim)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = rm, y = medv, color = crim), alpha=0.75) +
    geom_smooth(mapping = aes(x = rm, y = medv)) +
    coord_cartesian(xlim = c(4.5, 7.5)) +
    scale_y_log10() +
    scale_color_gradient(low = "yellow", high = "blue") +
    labs(x = "average rooms / house", y = "median house price ($K)",
       title = "Boston median house prices, house size, and crime")
```

Now color by `rad` but change all 24's to NA's.
```{r price-rad-na}
boston %>% 
  mutate(rad = ifelse(rad == 24, NA, rad)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = rm, y = medv, color = rad), alpha=0.75) +
    geom_smooth(mapping = aes(x = rm, y = medv)) +
    coord_cartesian(xlim = c(4.5, 7.5)) +
    scale_y_log10() +
    scale_color_gradient(low = "yellow", high = "red", na.value = "black") +
    labs(x = "average rooms / house", y = "median house price ($K)",
         title = "Boston median house prices and access to radial highways")
```

Maybe excluding newly-NA'ed `rad` values helps the crime plot.
```{r price-rooms-4}
boston %>% 
  filter(!rad == 24) %>% 
  mutate(crim = cume_dist(crim)) %>% 
  ggplot() +
    geom_point(mapping = aes(x = rm, y = medv, color = crim), size = 1) +
    geom_smooth(mapping = aes(x = rm, y = medv), lwd = 0.5) +
    scale_y_log10() +
    scale_color_gradient(low = "yellow", high = "blue") +
    labs(x = "average rooms / house", y = "median house price ($K)",
       title = "Boston median house prices, house size, and crime")
```

A grid of `nox` vs. `dis` plots according to `chas` (rows) and 
binned level (ntile) of `rad`.
```{r facet-grid, warning=F}
boston %>%
  mutate(rad = ifelse(rad == 24, NA, rad)) %>% 
  filter(!is.na(rad)) %>% 
  ggplot(aes(nox, dis, color = medv)) +
    geom_jitter() +
    facet_grid(chas ~ ntile(rad, 3)) +
    geom_smooth()
```

Multiplots available with `gridExtra`, used by ggplot2.
```{r multiple-plots}
require(gridExtra)
p1 <- ggplot(boston) +
  geom_point(aes(nox, medv, shape = chas, alpha = cume_dist(lstat)), 
             color = 'red', size = 2) +
  labs(title = 'using aes(alpha, shape)')
p2 <- boston %>%
  mutate(lstat_gr = ntile(lstat, 3)) %>% 
  ggplot(aes(tax, medv, color = lstat_gr, size = nox)) + 
  geom_point(shape = 16, alpha = 0.75) +
  geom_smooth(aes(group = lstat_gr), lwd = 0.8) +
  labs(title = 'using aes(group, size)')
p1
p2
ggsave('plots/two-plot.pdf', arrangeGrob(p1, p2))
```
