# dplyr and tidyr {#ch:dplyr}

```{r load-lahman}
library(tidyverse)
library(gridExtra)
batting <- as_tibble(Lahman::Batting)
fielding <- as_tibble(Lahman::Fielding)
```

## Hoofin' it with dplyr
Condense batting stats into player career totals, keep only those >= 500 games.
```{r gt-500-games}
is_col <- names(batting)[c(1, 2, 4, 6:17)]
is_num <- names(batting)[sapply(batting, is.numeric)]
gt_500 <- batting %>%
  select(is_col) %>%
  select(-teamID) %>% 
  drop_na() %>%
  group_by(playerID) %>% 
  summarize_at(is_col[-(1:3)], sum, na.rm = T) %>% 
  filter(G >= 500)
```

All Ha~ Green~ statistics to confirm that the data reduction looks right:
```{r show-greenha}
batting %>%
  filter(str_detect(playerID, "greenha"))  # a taste of `stringr`
```

Positions by game.
```{r count-pos}
fielding %>% 
  group_by(POS) %>% 
  count(wt = G)
```

Attach a column denoting their main fielding position.
```{r find-na-problem}
is_field = names(fielding)[c(1, 6, 7, 9, 10, 11, 12, 13)]
fielding %>%
  select(is_field) %>% 
  map(~ sum(is.na(.)))
```
That's odd, just one error NA.
```{r}
fielding %>% 
  filter(is.na(E))
```

Removing InnOuts is a good idea, too many missing,
and those `NA`s aren't relevant to the analysis.
```{r get-main-pos}
is_field = names(fielding)[c(1, 6, 7, 10, 11, 12, 13)]
pos_tot <- fielding %>%
  select(is_field) %>%   # cull columns
  drop_na() %>%  # drop the missing value
  group_by(playerID, POS) %>%  # want the most G by POS assigned to playerID
  summarize_all(sum) %>%
  ungroup() %>% 
  filter(G >= 100) %>%  # only those with 100 G at a POS
  arrange(playerID, desc(G)) %>%  # if G instead of desc(G), use last(POS)
  group_by(playerID) %>% 
  mutate(pos1 = first(POS)) %>% 
  filter(POS == pos1) %>%  # assign position with most games to POS
  select(-pos1)
```

## tidyr and relational data
Add fielding info to batting tibble.
```{r join-tibbles}
(batpos <- gt_500 %>% 
   left_join(pos_tot, by = "playerID", suffix = c(".h", ".f")))
```

Counts of positions.
```{r pos-player-totals}
batpos %>% 
  group_by(POS) %>% 
  count()
```

NAs are likely DHs.
```{r explore-na}
pos_nas <- batpos %>% 
  filter(is.na(POS))
batting %>% 
  inner_join(pos_nas, by = "playerID")
```

Drop these two DHs.
```{r drop-dh}
batpos <- batpos %>% 
  drop_na()
```

Now we could explore many aspects of hitting stats vs. position,
and see what position players were better fielders or better hitters,
or if neither we can see if they played for the Expos.
```{r ss-stats}
batpos %>% 
  filter(POS == "SS") %>% 
  mutate(BA = H / AB) %>%  # batting average, hits / at bats
  mutate(Err = E / (PO + A)) %>%   # error rate, errors / (put outs + assists)
  mutate(HRR = HR / AB) %>%  # home run rate, home runs / at bats
  ggplot(aes(Err, BA)) +
    geom_point(aes(color = HRR)) +
    geom_smooth()
```
```{r plot-pos-ba}
temp <- batpos %>% 
  mutate(BA = H / AB) %>%  # batting average, hits / at bats
  filter(between(BA, 0.01, 0.49)) %>% 
  mutate(Err = E / (PO + A)) %>%   # error rate, errors / (put outs + assists)
  mutate(HRR = HR / AB)  # home run rate, home runs / at bats
p1 <- temp %>% 
  ggplot(aes(Err, BA, color = POS)) +
    geom_point(alpha = 0.5, size = 0.5) +
    geom_smooth(aes(group = POS)) +
    coord_cartesian(xlim = c(0, 0.1), ylim = c(0.1, 0.42))
p2 <- temp %>% 
  filter(POS != "P") %>% 
  ggplot(aes(BA, HRR, color = POS)) +
    geom_point(alpha = 0.5, size = 0.5) +
    geom_smooth(aes(group = POS))
p1
p2
ggsave('plots/pos-bat.pdf', arrangeGrob(p1, p2))
```
