# dplyr and tidyr {#ch:dplyr}

```{r load-lahman}
library(tidyverse)
library(gridExtra)
batting <- as_tibble(Lahman::Batting)
fielding <- as_tibble(Lahman::Fielding)
```

## Hoofin' it with dplyr
Condense batting stats into player career totals, keep only those >= 500 games.
```{r gt-500-games}
is_col <- names(batting)[c(1, 2, 4, 6:17)]
is_num <- names(batting)[sapply(batting, is.numeric)]
gt_500 <- batting %>%
  select(is_col) %>%
  select(-teamID) %>% 
  drop_na() %>%
  group_by(playerID) %>% 
  summarize_at(is_col[-(1:3)], sum, na.rm = T) %>% 
  filter(G >= 500)
```

All Ha~ Green~ statistics:
```{r show-greenha}
batting %>%
  filter(str_detect(playerID, "greenha"))
```

Positions by game.
```{r count-pos}
fielding %>% 
  group_by(POS) %>% 
  count(wt = G)
```

Attach a column denoting their main fielding position.
```{r find-na-problem}
is_field = names(fielding)[c(1, 6, 7, 9, 10, 11, 12, 13)]
fielding %>%
  select(is_field) %>% 
  map(~ sum(is.na(.)))
```

Removing InnOuts is a good idea, too many missing.
```{r get-main-pos}
is_field = names(fielding)[c(1, 6, 7, 10, 11, 12, 13)]
pos_tot <- fielding %>%
  select(is_field) %>% 
  drop_na() %>% 
  group_by(playerID, POS) %>%
  summarize_all(sum) %>%
  ungroup() %>% 
  filter(G >= 100) %>% 
  arrange(playerID, desc(G)) %>%
  group_by(playerID) %>% 
  mutate(pos1 = first(POS)) %>% 
  filter(POS == pos1) %>% 
  select(-pos1)
```

## tidyr and relational data
Add fielding info to batting tibble.
```{r join-tibbles}
(batpos <- gt_500 %>% 
   left_join(pos_tot, by = "playerID"))
```

Counts of positions.
```{r pos-player-totals}
batpos %>% 
  group_by(POS) %>% 
  count()
```

NAs are likely DHs.
```{r explore-na}
pos_nas <- batpos %>% 
  filter(is.na(POS))
batting %>% 
  inner_join(pos_nas, by = "playerID")
```

Drop these two DHs.
```{r drop-dh}
batpos <- batpos %>% 
  drop_na()
```

Now we could explore many aspects of hitting stats vs. position,
and see what position players were better fielders or better hitters,
or if neither we can see if they played for the Expos.
```{r ss-stats}
batpos %>% 
  filter(POS == "SS") %>% 
  mutate(BA = H / AB) %>%  # batting average, hits / at bats
  mutate(Err = E / (PO + A)) %>%   # error rate, errors / (put outs + assists)
  mutate(HRR = HR / AB) %>%  # home run rate, home runs / at bats
  ggplot(aes(Err, BA)) +
    geom_point(aes(color = HRR)) +
    geom_smooth()
```
```{r plot-pos-ba}
temp <- batpos %>% 
  mutate(BA = H / AB) %>%  # batting average, hits / at bats
  filter(between(BA, 0.01, 0.49)) %>% 
  mutate(Err = E / (PO + A)) %>%   # error rate, errors / (put outs + assists)
  mutate(HRR = HR / AB)  # home run rate, home runs / at bats
p1 <- temp %>% 
  ggplot(aes(Err, BA, color = POS)) +
    geom_point(alpha = 0.5, size = 0.5) +
    geom_smooth(aes(group = POS)) +
    coord_cartesian(xlim = c(0, 0.1), ylim = c(0.1, 0.42))
p2 <- temp %>% 
  filter(POS != "P") %>% 
  ggplot(aes(BA, HRR, color = POS)) +
    geom_point(alpha = 0.5, size = 0.5) +
    geom_smooth(aes(group = POS))
p1
p2
ggsave('plots/pos-bat.pdf', arrangeGrob(p1, p2))
```
