--- 
title: "Hitchhiker's Guide to the Tidyverse (and Statistical Learning in R)"
author: "Cory Lanker"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [../clanker.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: clanker/tidyverse-class
description: "A quick exploration of tidyverse capabilities and implementation of statistical learning methodology for current R programmers"
---

# Introduction {-}

This bookdown notebook can be cloned via
```{bash git-clone, include=TRUE, eval=FALSE}
git clone git@github.com:clanker/tidyverse-class.git
```
<!--or downloaded as a 
#<zip file>[link](https://github.com/clanker/tidyverse-class/archive/master.zip)
-->

Introducing the tidyverse analyzing these data sets:

1. Basic plots with `tibble` and `ggplot2` using `Boston` house prices.
1. Preprocessing with `tidyr` and `dplyr` using `Lahman` baseball data.

**Other useful packages**

Though some of these commands will be used, we won't go deeply into
the following tidyverse packages. These packages
have an obvious function space, so knowing when to use
these packages and how to find the appropriate function
is easier than the packages discussed here.

1. Reading in data with `readr`.
1. String manipulation with `stringr`.
1. Dates and times with `lubridate`.
1. Handling factors with `forcats`.
1. Apply functions with `purrr`.

Some good ways to learn about these packages:

* `vignette()`, and search for documentation of that package,
* the cheat sheets for the packages on the RStudio website, or
* example("function") for helpful guidance on usage.

R proficiency is assumed. These notes aim to bring
a functional R coder into the tidyverse realm for modern
data analysis.

```{r install-package, eval=FALSE}
# To install the necessary packages in the tidyverse:
install.packages("tidyverse", dependencies = TRUE)
```

**to do list**

1. Add Chapter: computing using `caret`.
1. Add Chapter: functions provided by `mlr`.
1. Add Chapter: implementing `keras`.
1. Add Chapter: tips and tricks for better `R` coding.




```{r make-r-ref, include=FALSE, eval=FALSE} 
# only have to eval once
# automatically create a bib database for R packages
knitr::write_bib(c(.packages(), 'bookdown', 'knitr', 'rmarkdown',
                   'ggplot2', 'dplyr', 'tidyr', 'rlang',
                   'caret', 'mlr', 'sjmisc'), 'packages.bib')
```

Many references are made to Hadley Wickham's book, 
_R for Data Science_ [@wickham2016r].
This document is built with R Markdown, **knitr** [@xie2015],
and the **bookdown** package [@R-bookdown].
