--- 
title: "Hitchhiker's Guide to the Tidyverse"
author: "Cory Lanker"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [../clanker.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: clanker/tidyverse-class
description: "A quick exploration of tidyverse capabilities for current R programmers, heavily based on Hadley Wickham's notes from _R for Data Science_"
---

# Introduction {-}

<!--
This bookdown notebook can be cloned via
```{bash git-clone, include=TRUE, eval=FALSE}
git clone git@github.com:clanker/tidyverse-class.git
```
or downloaded as a 
<zip file>[link](https://github.com/clanker/tidyverse-class/archive/master.zip).
-->

Introducing the tidyverse with two data sets:

1. Basic plots with `tibble` and `ggplot2` using `Boston` house prices.
1. Preprocessing with `tidyr` and `dplyr` using `Lahman` baseball data.

Though some of these commands will be used, we won't go deeply into
the following tidyverse packages:

1. Reading in data with `readr`.
1. String manipulation with `stringr`.
1. Dates and times with `lubridate`.
1. Handling factors with `forcats`.

R proficiency is assumed. These notes aim to bring
a functional R coder into the tidyverse realm.

```{r install-package, eval=FALSE}
# To install the necessary packages in the tidyverse:
install.packages("tidyverse", dependencies = TRUE)
```


```{r make-r-ref, include=FALSE, eval=FALSE} 
# only have to eval once
# automatically create a bib database for R packages
knitr::write_bib(c(.packages(), 'bookdown', 'knitr', 'rmarkdown'), 'packages.bib')
```

This document is built with R Markdown, **knitr** [@xie2015],
and the **bookdown** package [@R-bookdown].
